The Evolution of Phenotypically Invariant Genetic Networks

### 2018-03-09 kalman_walk_paper.tex
added some details

\paragraph{Hybridization}
The means of two allopatric populations each of effective size $N_e$ separated for $T$ generations
will be a distance roughly of order $2\sigma_N \sqrt{T/N_e}$ apart along $\optx$.
(Consult figure \ref{fig:conceptual_fig} for a conceptual diagram.)
A population of $F_1$ hybrids has one haploid genome from each,
whose coefficients are averaged,
and so will have mean system coefficients at the midpoint between their means.
The distribution of $F_2$ hybrids will have mean at the average of the two populations,
but will have higher variance \citep{barton2001role,rosas2010cryptic,chevin2014niche}.
This ``segregation variance'' of $F_2$ hybrids 
can be shown to increase linearly with the square of the distance between
parental population means
under models of both simple and polygenic traits.
This is suggested by figure \ref{fig:conceptual_fig} and shown by \citet{slatkin1994segregation}
(also see Appendix \ref{apx:seg_cov} for a different derivation).
Concretely, we expect the population of $F_1$s to have variance $\sigma^2_S$ in the selected direction
(the same as within each parental population),
but the population of $F_2$ hybrids will have variance of order $\sigma^2_S + 4 \omega \sigma^2_N T/N_e$,
where $\omega$ is a factor that depends on the genetic basis of the coefficients.
If the optimal set $\allS$ has dimension $q$,
using the polygenic model of appendix \ref{apx:seg_cov}, 
$\omega$ is proportional to the number of degrees of freedom: $\omega = (p-q)/8$.
If each trait is controlled by a single locus, as in figure \ref{fig:conceptual_fig},
the value is similar.

@@

Substantial analytical work has made similar observations
using Fisher's geometric model, 
including Haldane's rule and the $F_2$ ``speciation clock'' 
\citep{barton2001role,fraisse2016genetics,simon2017coadapted}.
These calculations are similar in spirit, 
but differ from our own in that they primarily assume a monomorphic population.
Indeed, our model may provide an additional mechanistic context 
in which Fisher's geometric model is appropriate,
although the typical degree of pleiotropy
and stability of the $G$-matrix are unknown in practice.
Our observations argue for the importance of including neutral directions in these models,
which is not usually done (but see \citet{chevin2014niche}).

@@

\paragraph{Fisher's geometric model}
There has been substantial work on models of quadratic stabilizing selection
around a single optimum -- i.e., Fisher's geometric model \citep[e.g.][]{simon2017coadapted}.
Our model would have this form if viewed only in phenotype space, 
where there is a single optimum --
but differs by explicitly considering evolutionary dynamics of underlying genotype-derived regulatory matrices.
These make the quantitative genetic model we are led to
a generalization of Fisher's geometric model,
with distinct behavior,
due to the existence of neutral directions in genotype space.
However, there are many shared aspects:
for instance, \citet{barton1989divergence} also found incompatibilities accumulating
with genetic drift.
Fisher's geometric model also predicts Haldane's rule
and many other empirically observed patterns \citep{fraisse2016genetics},
but for different underlying reasons.

@@

Suppose that a population has a Gaussian distribution in $d$-dimensional trait space
with mean $\mu$ and covariance matrix $\Sigma$,
and that fitness of an individual at $x$ is $\exp(-\|Lx\|^2/2)$.
Then, completing the square as above with $A=\Sigma^{-1}$, $y=\mu$, and $B=L^T L$,
and defining $Q = (\Sigma^{-1} + L^T L)^{-1}$,
\begin{align*}
    &
    \frac{1}{\sqrt{2 \pi}^n \det(\Sigma)^{1/2}} 
        \int e^{-\frac{1}{2} x^T \Sigma^{-1} x} e^{-\frac{1}{2} x^T L^T L x} dx \\
    &\qquad =
    \frac{1}{\sqrt{2 \pi}^n \det(\Sigma)^{1/2}} 
        \int e^{-\frac{1}{2}(x-Q \Sigma^{-1} \mu)^T Q^{-1} (x-Q \Sigma^{-1} \mu)} dx 
        \\ &\qquad \qquad {}
        \times e^{\mu^T \left( I - \Sigma^{-1} Q \right) \Sigma^{-1} \mu} \\
    &\qquad =
    \sqrt{\frac{\det(Q)}{\det(\Sigma)}}
        \exp\left\{\mu^T \left( I - \Sigma^{-1} Q \right) \Sigma^{-1} \mu\right\} \\
    &\qquad =
    \sqrt{\frac{1}{\det(\Sigma)\det(\Sigma^{-1}+L^TL)}}
        \exp\left\{\mu^T \left( I - (I + L^T L \Sigma)^{-1} \right) \Sigma^{-1} \mu\right\} \\
\end{align*}

@@

The description above does not describe how \emph{two} diverging populations interact,
since the amount of \emph{segregation variance}, quantified by $R$,
will not stay constant.
To get an idea of how this might change,
suppose that a trait is determined by $L$ unlinked, biallelic loci,
and that the $i^\mathrm{th}$ locus has two alleles with additive effects $\pm \alpha_i$,
so that being homozygous for the $+$ allele contributes $+2 \alpha_i$ to the trait.
For simplicity, we will neglect the effects of selection.
If the $+$ allele at locus $i$ is at frequency $p_i$ in a population,
then the mean and genetic variance of the trait in a diploid population with random mating is
\begin{align*}
    m &= 2 \sum_i (2 p_i - 1) \alpha_i \\
    s^2 &= 4 \sum_i p_i (1-p_i) \alpha_i^2 .
\end{align*}

>>>>

\paragraph{Hybridization}
The means of two allopatric populations each of effective size $N_e$ separated for $T$ generations
will be a distance roughly of order $2\sigma_N \sqrt{T/N_e}$ apart along $\optx$.
(Consult figure \ref{fig:conceptual_fig} for a conceptual diagram.)
A population of $F_1$ hybrids has one haploid genome from each,
whose coefficients are averaged,
and so will have mean system coefficients at the midpoint between their means.
The distribution of $F_2$ hybrids will have mean at the average of the two populations,
but will have higher variance \citep{barton2001role,rosas2010cryptic,chevin2014niche}.
This ``segregation variance'' of $F_2$ hybrids 
can be shown to increase linearly with the square of the distance between
parental population means
under models of both simple and polygenic traits.
This is suggested by figure \ref{fig:conceptual_fig} and shown by \citet{slatkin1994segregation}
(also see Appendix \ref{apx:seg_cov} for a different derivation).
Concretely, we expect the population of $F_1$s to have variance $\sigma^2_S$ in the selected direction
(the same as within each parental population),
but the population of $F_2$ hybrids will have variance of order $\sigma^2_S + 4 \omega \sigma^2_N T/N_e$,
where $\omega$ is a factor that depends on the genetic basis of the coefficients
\citep[the ``segregation variance''][]{slatkin1994segregation}.
If the optimal set $\allS$ has dimension $q$,
using the polygenic model of appendix \ref{apx:seg_cov}, 
$\omega$ is proportional to the number of degrees of freedom: $\omega = (p-q)/8$.
If each trait is controlled by a single locus, as in figure \ref{fig:conceptual_fig},
the value is similar.

@@

%% Fisher's geometric model
% Differences:
% - they look at shifting environments too
% - they just model substitutions, so only hybrids have genetic variance
% chevin2014niche:
% - m is accessible, constrained dimensionlity (allows for a ridge)
% - predicts rate of RI accumulation (due to substitutions) is \propto N^{-(1+m/2)}
%   since this is mean log fitnees; says hybrids will have fitness
%     exp( - m C T N^{-(1+m/2)} ) ~ 1 - m C T N^(-(1+m/2))
% - says RI accum is *faster* at higher m but this is not obv frm eqns
% - we say it should be ( exp(-C (T/N)^2) / sqrt(T/N) )^m :
%     ~ 1 - m C (T/N)

\paragraph{Fisher's geometric model}
Substantial analytical work on speciation has been done using Fisher's geometric model, 
which can also predict Haldane's rule and the $F_2$ ``speciation clock'' 
\citep{barton2001role,fraisse2016genetics,simon2017coadapted}.
The model is similar to the quantitative genetics model we use 
to predict the rate of speciation, 
although work on Fisher's geometric model generally only models substitutions
between populations, i.e., neglecting within-species polymorphism.
Indeed, our model may provide an additional mechanistic context 
in which Fisher's geometric model is appropriate,
although the typical degree of pleiotropy
and stability of the $G$-matrix are unknown in practice.
Our observations argue for the importance of including neutral directions in these models,
which is not usually done (but see \citet{chevin2014niche}).

Although our main focus is on the implications of system theory,
rather than on the quantitative genetics modeling,
it is interesting to compare analytic results.
\citet{chevin2014niche} finds that mean fitness decrease of hybrids
after $T$ generations grows proportionally to $m N_e^{-1-m/2} T$, for large $N_e$;
while our corresponding rate is $m T / N_e$.
This difference in time scaling is substantial, especially for large populations,
and seems to appear because \citet{chevin2014niche} uses the results of \citet{sella2005application}
to describe how the steady-state fitness and substitution process
depends on trait dimensionality
(the fitness peak is harder to find in higher dimensions).
We do not incorporate this effect
because genetic variation may be provided by other sources,
but if we did, it would enter through our $\sigma_S$.

@@

@@

Suppose that a population has a Gaussian distribution in $d$-dimensional trait space
with mean $\mu$ and covariance matrix $\Sigma$,
and that fitness of an individual at $x$ is $\exp(-\|Lx\|^2/2)$.
Then, completing the square as above with $A=\Sigma^{-1}$, $y=\mu$, and $B=L^T L$,
and defining $Q = (\Sigma^{-1} + L^T L)^{-1}$,
\begin{align*}
    &
    \frac{1}{\sqrt{2 \pi}^n \det(\Sigma)^{1/2}} 
        \int e^{-\frac{1}{2} x^T \Sigma^{-1} x} e^{-\frac{1}{2} x^T L^T L x} dx \\
    &\qquad =
    \frac{1}{\sqrt{2 \pi}^n \det(\Sigma)^{1/2}} 
        \int e^{-\frac{1}{2}(x-Q \Sigma^{-1} \mu)^T Q^{-1} (x-Q \Sigma^{-1} \mu)} dx 
        \\ &\qquad \qquad {}
        \times e^{\mu^T \left( I - \Sigma^{-1} Q \right) \Sigma^{-1} \mu} \\
    &\qquad =
    \sqrt{\frac{\det(Q)}{\det(\Sigma)}}
        \exp\left\{\mu^T \left( I - \Sigma^{-1} Q \right) \Sigma^{-1} \mu\right\} \\
    &\qquad =
    \sqrt{\frac{1}{\det(\Sigma)\det(\Sigma^{-1}+L^TL)}}
        \exp\left\{\mu^T \left( I - (I + L^T L \Sigma)^{-1} \right) \Sigma^{-1} \mu\right\} \\
\end{align*}
This is the product of two terms:
the first gives the drop in fitness due to segregation variance,
and the second the drop due to a shift in mean away from the optimum.
A similar decomposition appears in \citet{barton2001role} and \citet{chevin2014niche},
but for mean log fitness.

@@

The description above does not describe how \emph{two} diverging populations interact,
since the amount of \emph{segregation variance}, quantified by $R$,
will not stay constant.
This has been quantified in various models before
\citep[e.g.,][]{slatkin1994segregation,barton2001role,chevin2014niche};
we provide the following calculations for completeness.

To get an idea of how segregation variance might evolve,
suppose that a trait is determined by $L$ unlinked, biallelic loci,
and that the $i^\mathrm{th}$ locus has two alleles with additive effects $\pm \alpha_i$,
so that being homozygous for the $+$ allele contributes $+2 \alpha_i$ to the trait.
For simplicity, we will neglect the effects of selection.
If the $+$ allele at locus $i$ is at frequency $p_i$ in a population,
then the mean and genetic variance of the trait in a diploid population with random mating is
\begin{align*}
    m &= 2 \sum_i (2 p_i - 1) \alpha_i \\
    s^2 &= 4 \sum_i p_i (1-p_i) \alpha_i^2 .
\end{align*}
